---
title: "01-Cleaning Faaborg et al Guanica Forest migrant data"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Data originally from Table 2 in Faaborg et al 2013.

Tables in "wide" format and need to be cleaned and reshaped to "long" format for analysis in R

### References
Faaborg J, WJ Arendt, JD Toms, KM Dugger, WA Cox, and  M Canals Mora.  2013. Long-term decline of a winter-resident bird community in Puerto Rico.  Biodiversity & Conservation.  https://link.springer.com/article/10.1007/s10531-012-0399-7
Table 2: "Number of winter residents captured annually from 1973–2012"


Long‐term variation in the winter resident bird community of Guánica Forest, Puerto Rico: lessons for measuring and monitoring species richness
J Faaborg, KM Dugger, WJ Arendt. - Journal of Field Ornithology, 2007.   http://www.jstor.org/stable/40345963
Table 1: data from 1989 through 2006.

Understanding survival and abundance of overwintering warblers: does rainfall matter?
KM Dugger, J Faaborg, WJ Arendt, KA Hobson - The Condor, 2004 
http://www.bioone.org/doi/abs/10.1650/7632 ; available on JSTOR
Summarizes environmental variables in Table 1


### Steps

* Load .csv with wide data (Table 2, Faabor et al 2013)
* Clean data (eg, convert species names to AOU codes)
* Reshape data to long format using reshape2::melt

### Analysis History / Original file names:

* Original analysis carried out by NLB/EC in 2017
* Turned into research compendium in 6/2017
* Original file scripts began with was "FaaborgData_edit_CSV.csv"
  - flagged as "edit" because... ?
  - file chanaged to "faaborg_2013.csv"
  
  
## Libraries

```{r}
library(reshape2)
library(here)
```


## Load Faaborg et al 2013 data

Note that the location of these data depends on whether

* You downloaded the data and saved it to your working directory
  - Set your working directory or the file path accordinly
* You loaded faanica into R as a package
  - the file path should be "./extdata/faaborg2013.csv"
* You forked or cloned the faanica github repro (or are the maintainer) and are therefore working from the faanica sources code
  -  ".inst/extdata/faaborg2013.csv"

I am working directly from the faanica source code and will load from "".inst/extdata/faaborg2013.csv""

Use here() function in here package to assure that file name always gets properly constructed (can be a problem in some contexts)
```{r}
file. <- here::here("inst/extdata/faaborg2013_table2.csv")
```

Load the data; note that there is a 9-row header of meta data that needs to be skipped
```{r}
faa.wide <- read.csv(file., comment.char = "#", skip = 9)
```

## Clean data

Set up 4 letter AOU species codes from common names.

### Create spp names
```{r}
#extract 1st 2 letters of each part of name
faa.wide$spp.code <- gsub("^([A-Za-z]{2})(.*)([ ])([A-Za-z]{2})(.*)","\\1\\4", faa.wide$common.name)

## Change spp.code to upper case
faa.wide$spp.code <- toupper(faa.wide$spp.code)

## Correct species names that don't follow pattern from above

## Merlin
faa.wide$spp.code[which(faa.wide$spp.code == "MERLIN")] <- "MERL"

## Ovenbird
faa.wide$spp.code[which(faa.wide$spp.code == "OVENBIRD")] <- "OVEN"

## Worm-eating warbler
faa.wide$spp.code[which(faa.wide$spp.code == "WOWA")] <- "WEWA"

## Distinguish blue-winged warbler (BWWA), black-and-white warbler (BAWW), and black-throated blue warbler (BTBW)
faa.wide[which(faa.wide$spp.code == "BLWA"), c("spp.code","common.name")] 
faa.wide$spp.code[which(faa.wide$spp.code == "BLWA")] <- c("BWWA" ,"BAWW","BTBW")

## Distinguish prairie warbler (PRWA) and prothonotary warbler (PROW)
faa.wide[which(faa.wide$spp.code == "PRWA"),c("spp.code","spp")] 
faa.wide$spp.code[which(faa.wide$spp.code == "PRWA")][[1]] <- "PROW"

## Cape May warbler (CMWA)
faa.wide$spp.code[which(faa.wide$spp.code == "CAWA")] <- "CMWA"

## Rose-breasted grosbeak (RBGR)
faa.wide$spp.code[which(faa.wide$spp.code == "ROGR")] <- "RBGR"
```

### Validate species codes

Look at summary of species codes to make sure there is only one instance of each
```{r}
with(faa.wide, 
     table(spp.code))
```

Use head() and tail() to look at first and last rows of dataframe, Confirm nothing has been added or deleted, Make sure everything looks as expected.
```{r}
head(faa.wide)
tail(faa.wide)
```

## Convert from wide- to long-format & clean 


### Melt data
* ID variables uniquely identify each row of data
  - common.name, spp, spp.code
* specify that you want "year" and "N" that correspond to each (common.name, spp, spp.code)

```{r}
dat.melt <- reshape2::melt(data = faa.wide,
     id.vars = c("common.name",
                 "spp",
                 "spp.code"),
     variable.name = "year",
     value.name = "N")
```


### Clean melted data

* Remove "yr." from in front of the year
* Tell R that "year" is numeric
```{r}
dat.melt$year <- gsub("yr.", "", dat.melt$year)
dat.melt$year <- as.numeric(dat.melt$year)
```


## Append data on number of nets per year

* The following code shows how to solve this problem using loops
* the merge() function can also do this

### Add number of nets used each year 

* ... prior to 1989?
* 7 in 1989
* 8 in 1990
* 9 in all subsequent years
```{r}
n.rows <- nrow(dat.melt)
dat.melt$nets <- rep(9,n.rows)

for (i in 1:480)
{
    if ((dat.melt$year[i] == 1989))
    {
      dat.melt$nets[i] <- 7
    }
}

for (i in 1:480)
{
    if ((dat.melt$year[i] == 1990)) 
    {
      dat.melt$nets[i] <- 8
    }
}
```



## Calculate caps per net

Mist ned dat is frequently expressed as capture per net hour.

```{r}
dat.melt$caps.per.net <- with(dat.melt, N/nets)
```


## Save long data

```{r}
file. <- here::here("inst/extdata/faaborg2013.csv")
write.csv(dat.melt,
          file = file.,
          row.names = F)
```


```{r eval=FALSE, include=FALSE}
#save data as R object to /dat
migrants <- dat.melt
use_data(migrants)
```

